Synthesizing Symmetric Lenses

### 2018-07-11 papers/SymmetricLenses/main.tex
minor edit
deleted extra paragraph in section 2
big edit on related

Two classical symmetric lenses are equivalent if they output the same
formats given any sequence of edits. Formally, given a lens
$\Lens \in X \leftrightarrow Y$,
an \emph{edit} for $\Lens$ is a member of $X + Y$. Consider
the function $apply$, which, given a lens and an element of that lens's
complement, is a function from sequences of edits to sequences of edits. If
$apply(\Lens,c,es) = es'$, then that means, starting with complement $c$, after
edit $es_i$, the lens $\Lens$ synchronizes the other data to $es'_i$.
\begin{centermath}
\begin{tabular}[b]{l@{\qquad \; \qquad}l}
$
  \inferrule
  {
  }
  {
    apply(\Lens,c,[]) = []
  }
$
&
$
  \inferrule
  {
    \Lens.putr(x,c) = (y,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InLOf{x})::es) = (\InROf{y})::es'
  }
$
\end{tabular}
\end{centermath}
\[
  \inferrule
  {
    \Lens.putl(y,c) = (x,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InROf{y})::es) = (\InLOf{x})::es'
  }
\]
\noindent
Two lenses, $\Lens_1$ and $\Lens_2$, are equivalent if
$apply(\Lens_1,\Lens_1.init,es) = apply(\Lens_2,\Lens_2.init,es)$ for all edit
sequences $es$.

@@

%\paragraph*{Issues with Classical Symmetric Lenses}
\iffalse
Using symmetric lenses for
synchronization forces us to maintain not just two files that we aim to
sycnrhonize, but three. \bcp{I don't find it unintuitive, and I think we
  should be careful not to oversell the difficulty of storing the
  complements.  It requires more engineering, but it's not rocket
  science.}This is unintuitive, and creates many problems in 
practice. For example, consider the situation where I am synchronizing a crontab
and job manager file. I am synchronizing across computers, so I must either
maintain a copy of the complement on each computer, and keep them synchronized,
or figure out a computer to hold the complement, and make sure all edits go
through that complement. Furthermore, if I want to change the synchronization to
a different computer, I don't need to merely make sure that my files are moved
to the computer, I also have to make sure that I correctly move the complement,
and hook up that complement to be used for that synchronization task. This is a
huge divergence from other synchronization utilities\bcp{No.  For example,
  Unison maintains a database on each replica recording, for each file, what
its hash was after the last time the two replicas were synchronized.}, where all
that needs to be 
done is maintain the files. Using those for synchronization, edits can be sent
over the wire, and the appropriate function can stitch together those edits, no
additional state needed.
\fi
While classical symmetric are more expressive than simple symmetric
lenses, it is unclear whether this additional expressiveness is useful in
practice.  Moreover, classic symmetric lenses come with 
some disadvantages.
For one, an implementation of classical symmetric lenses
would need to squirrel away the complement in
an auxiliary file somwhere, which likely requires more engineering.
However, more significant issues arise in synthesis.
First, because each lens has a custom complement, one can no longer
specify the put functions through input/output examples alone.  One
alternative would be to enrich specifications with edit sequentials;
another would be to specify the structure of complements explicitly
(though the latter would be somewhat akin to specifying the internal
state of a program).  In either case, the complexity of the specifications
increases.
%% addition, users would have to specify some auxilary 
%% to specify the behavior through the observable behavior of edit sequences.
%% Second, there are lenses that require arbitrarily long edit sequences to
%% differentiate between: consider the lens that keeps two formats equal for the
%% first $k$ non-identity edits, then keeps them completely unsynchronized for all
%% future edits.  This requires an edit sequence of $k$ edits to differentiate
%% between this lens and the identity lens.  Simple symmetric lenses avoid this
%% problem by disallowing such lenses: we can't keep track of the number of
%% non-identity edits performed unless there are explicit fields in the data
%% formats keeping track of them.  While we are reducing the expressivity of our
%% lenses with this restriction, we feel the benefits outweigh the costs.

\paragraph*{Relation with Simple Symmetric Lenses} To formalize how simple
symmetric lenses work as synchronization tools, we define an $apply$ function on
them, much like one defines an $apply$ on classical symmetric lenses: Given a
simple symmetric lens and an optional pair of source and target strings, $apply$
is a function from sequences of edits to sequences of edits. If
$apply(\Lens,\None,es) = es'$, then starting with no prior data, after edit
$es_i$, the lens \Lens synchronizes the updates the data in the non-edited
format (the right format if $es_i = \InLOf{x}$, and the left format if $es_i =
\InROf{y}$) to $es_i'$. If $apply(\Lens,\SomeOf{(x,y)},es) = es'$, then that
means, starting with data $x$ and $y$ on the left and right, respectively, after
edit $es_i$, the lens \Lens synchronizes the other data to $es_i'$.

@@

We have reformulated symmetric lenses in a style similar to existing asymmetric
lens formulations. However, some symmetric lenses certainly are inexpressible in
the simple symmetric lens formulation. We formalize which are expressible
with a restriction on symmetric lenses. We define \emph{forgetful symmetric
  lenses} to be symmetric lenses that satisfy the following additional laws.
\begin{equation}
  \tag{\ForgetfulRL}
  \begin{mathprooftree}
    \AxiomC{$\Lens.putr(x,c_1) = (\_,c_1')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putr(x,c_2) = (\_,c_2')$}
    \AxiomC{$\Lens.putl(y,c_1') = (\_,c_1'')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putl(y,c_2') = (\_,c_2'')$}
    \def\extraVskip{2pt}
    \singleLine
    \BinaryInfC{$c_1'' = c_2''$}
  \end{mathprooftree}
\end{equation}
\begin{equation}
  \tag{\ForgetfulLR}
  \begin{mathprooftree}
    \AxiomC{$\Lens.putl(y,c_1) = (\_,c_1')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putl(y,c_2) = (\_,c_2')$}
    \AxiomC{$putr(x,c_1') = (\_,c_1'')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putr(x,c_2') = (\_,c_2'')$}
    \def\extraVskip{2pt}
    \singleLine
    \BinaryInfC{$c_1'' = c_2''$}
  \end{mathprooftree}
\end{equation}

Intuitively, these equations state that complements are uniquely determined by the most
recently input $x$ and $y$. Hence, such lenses correspond closely \dpw{replace ``closely'' with
``exactly''?} with simple symmetric
lenses, where all state is maintained by the current state of the $x$ and $y$
data. Forgetful symmetric lenses express exactly the same $apply$ function as
simple symmetric lenses.

@@

\bcp{I guess we could also show that equivalence of simple symmetric lenses
  is just the identity relation?}
\sam{No; for example the identity lens of a concatenation of two regular
  expressions is the same as concatenating the two identity lenses on the
  respective regular expressions}
\afm{I think Benjamin likely means identity relation on the semantic meaning of
  the lenses, which is true.  Would this be a useful thing to prove? It's a 2
  liner.}

@@

All pragmatic synthesis algorithms use heuristics of one kind or another.  One of the
goals of the current paper is to try to ground those heuristics in a broader theory,
information theory,
with the hope that this theory may inform future design decisions and help us understand
heuristics crafted in other tools and possibly in other domains.  For instance, we
speculate that some of the heuristics used in
FlashFill (to take one well-documented example) may be connected to some of the principles
laid out here.  For instance, FlashFill prioritizes the substring constructor over the
constant string constructor when ranking possible programs. Such a choice may be understood
in information-theoretic terms as the translation from target back to source would require
more information.  Said another way, our information theoretic metrics measure injectivity
(in both directions, in our case, to measure bijectivity) but applied in one direction, as
is the case for FlashFill, would avoid the highly injective constant functions.  Beyond
constants, another FlashFill heuristic prefers ``wider'' character classes (or token classes)
over ``narrower'' ones.  Selecting a wider set of strings to copy from left-to-right, viewed
in the reverse direction, from right-to-left, results in a function that is closer to being
injective, as measured by information theory.  These are just a few of the heuristics used
in Flashfill; more research would be required (and public disclosure of the details necessary)
to understand the connections fully.

@@

More broadly, information theory and probabilistic languages are common tools
in natural language understanding,  machine translation, information retrieval,
data extraction and grammatical inference (see Pereira~\cite{pereira:info-theory},
for an introduction).  Indeed, our work was inspired, in part, by the PADS format
inference tool~\cite{pads:synthesis}, which was in turn inspired by earlier work on
probabilistic gramatical inference
and information extraction~\cite{kushmerick-thesis,arasu:extracting}.  More specifically, PADS
used an information-theoretic metric to help judge the quality of inferred grammars
for ad hoc data formats.  However, PADS did not synthesize data transformers, and
we are not aware of the use of information-theoretic principles in type-directed or syntax-guided
synthesis of DSL programs.

@@

% begin conclusion
\section{Conclusion}
\label{sec:conc}
We have developed a synthesis algorithm for synthesizing synchronization
functions between data formats that may not be in bijective correspondence. We
do this by synthesizing symmetric lenses. However, existing symmetric lens
formulations use complex complements to store external state. We identify a
restriction on symmetric lenses, simple symmetric lenses. We develop combinators
for simple symmetric lenses on string data. We show how to synthesize these
lenses from regular expression specifications. Prior work could return
\emph{any} satisfying lens, as the type system was restrictive enough that any
satisfying lens would typically be the correct satisfying lens. Because of the
increased flexibility of the symmetric lens language, we identify an
information-theoretic heuristic for finding a likely satisfying lens. We
developed a greedy synthesis algorithm that aims to find the lens of lowest
cost. We demonstrate empiraclly that our algorithm can find lenses quickly, and
our heuristic makes finding the correct lens require fewer specifications.

@@

\begin{centermath}
\begin{array}[b]{l@{\qquad \; \qquad}l}
  \inferrule*
  {
    \String \in \LanguageOf{\BRegex}\\
    \StringAlt \in \LanguageOf{\BRegexAlt}
  }
  {
    \DisconnectOf{\BRegex}{\BRegexAlt}{\String}{\StringAlt}
    \OfType \BRegex \Leftrightarrow \BRegexAlt
  }
&
  \begin{array}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String'$ & = & $\StringAlt$\\
    $\CreateL{} \App \StringAlt'''''$ & = & $\String$\\
    $\PutR{} \App \String''' \App \StringAlt'$ & = & $\StringAlt'$\\
    $\PutL{} \App \StringAlt' \App \String'$ & = & $\String'$
  \end{array}
\end{array}
\end{centermath}

>>>>

Two classical symmetric lenses are equivalent if they output the same formats
given any sequence of edits. Formally, given a lens $\Lens \in X \leftrightarrow
Y$, an \emph{edit} for $\Lens$ is a member of $X + Y$. Consider the function
$apply$, which, given a lens and an element of that lens's complement, is a
function from sequences of edits to sequences of edits. If $apply(\Lens,c,es) =
es'$, then given complement $c$ and edits $es_i$, the lens $\Lens$ generates
$es'_i$.
\begin{centermath}
\begin{tabular}[b]{l@{\qquad \; \qquad}l}
$
  \inferrule
  {
  }
  {
    apply(\Lens,c,[]) = []
  }
$
&
$
  \inferrule
  {
    \Lens.putr(x,c) = (y,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InLOf{x})::es) = (\InROf{y})::es'
  }
$
\end{tabular}
\end{centermath}
\[
  \inferrule
  {
    \Lens.putl(y,c) = (x,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InROf{y})::es) = (\InLOf{x})::es'
  }
\]
\noindent
Two lenses, $\Lens_1$ and $\Lens_2$, are equivalent if
$apply(\Lens_1,\Lens_1.init,es) = apply(\Lens_2,\Lens_2.init,es)$ for all $es$.

@@

%\paragraph*{Issues with Classical Symmetric Lenses}
\iffalse Using symmetric lenses for synchronization forces us to maintain not
just two files that we aim to sycnrhonize, but three. \bcp{I don't find it
  unintuitive, and I think we should be careful not to oversell the difficulty
  of storing the complements. It requires more engineering, but it's not rocket
  science.}This is unintuitive, and creates many problems in practice. For
example, consider the situation where I am synchronizing a crontab and job
manager file. I am synchronizing across computers, so I must either maintain a
copy of the complement on each computer, and keep them synchronized, or figure
out a computer to hold the complement, and make sure all edits go through that
complement. Furthermore, if I want to change the synchronization to a different
computer, I don't need to merely make sure that my files are moved to the
computer, I also have to make sure that I correctly move the complement, and
hook up that complement to be used for that synchronization task. This is a huge
divergence from other synchronization utilities\bcp{No. For example, Unison
  maintains a database on each replica recording, for each file, what its hash
  was after the last time the two replicas were synchronized.}, where all that
needs to be done is maintain the files. Using those for synchronization, edits
can be sent over the wire, and the appropriate function can stitch together
those edits, no additional state needed. \fi While classical symmetric lenses
are technically more expressive than simple symmetric lenses, it is unclear
whether this additional expressiveness is useful in practice. In our judgement,
the complications they introduce into synthesis did not justify the potential
additional expressiveness: because each lens has a custom
complement, one can no longer specify the put functions through input/output
examples alone. One alternative would be to enrich specifications with edit
sequentials; another would be to specify the structure of complements explicitly
(though the latter would be somewhat akin to specifying the internal state of a
program). In either case, the complexity of the specifications increases.
%% addition, users would have to specify some auxilary 
%% to specify the behavior through the observable behavior of edit sequences.
%% Second, there are lenses that require arbitrarily long edit sequences to
%% differentiate between: consider the lens that keeps two formats equal for the
%% first $k$ non-identity edits, then keeps them completely unsynchronized for all
%% future edits.  This requires an edit sequence of $k$ edits to differentiate
%% between this lens and the identity lens.  Simple symmetric lenses avoid this
%% problem by disallowing such lenses: we can't keep track of the number of
%% non-identity edits performed unless there are explicit fields in the data
%% formats keeping track of them.  While we are reducing the expressivity of our
%% lenses with this restriction, we feel the benefits outweigh the costs.

\paragraph*{Relation with Simple Symmetric Lenses} To compare classic and simple
symmetric lenses, we define an $apply$ function on simple lenses as well. If
$apply(\Lens,\None,es) = es'$, then starting with no prior data, after edit
$es_i$, the lens \Lens generates $es_i'$ (the right format if $es_i =
\InLOf{x}$, and the left format if $es_i = \InROf{y}$). If
$apply(\Lens,\SomeOf{(x,y)},es) = es'$, then starting with data $x$
and $y$ on the left and right, respectively, after edit $es_i$, the lens \Lens
generates $es_i'$.

@@

Next, we define \emph{forgetful symmetric
  lenses} to be symmetric lenses that satisfy the following additional laws.
\begin{equation}
  \tag{\ForgetfulRL}
  \begin{mathprooftree}
    \AxiomC{$\Lens.putr(x,c_1) = (\_,c_1')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putr(x,c_2) = (\_,c_2')$}
    \AxiomC{$\Lens.putl(y,c_1') = (\_,c_1'')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putl(y,c_2') = (\_,c_2'')$}
    \def\extraVskip{2pt}
    \singleLine
    \BinaryInfC{$c_1'' = c_2''$}
  \end{mathprooftree}
\end{equation}
\begin{equation}
  \tag{\ForgetfulLR}
  \begin{mathprooftree}
    \AxiomC{$\Lens.putl(y,c_1) = (\_,c_1')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putl(y,c_2) = (\_,c_2')$}
    \AxiomC{$putr(x,c_1') = (\_,c_1'')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putr(x,c_2') = (\_,c_2'')$}
    \def\extraVskip{2pt}
    \singleLine
    \BinaryInfC{$c_1'' = c_2''$}
  \end{mathprooftree}
\end{equation}

Intuitively, these equations state that complements are uniquely determined by
the most recently input $x$ and $y$. Hence, such lenses correspond exactly with
simple symmetric lenses, where all state is maintained by the current state of
the $x$ and $y$ data. Forgetful symmetric lenses express exactly the same
$apply$ function as simple symmetric lenses.

@@

@@

All pragmatic synthesis algorithms use heuristics of one kind or another.  One of the
goals of the current paper is to try to ground those heuristics in a broader theory,
information theory,
with the hope that this theory may inform future design decisions and help us understand
heuristics crafted in other tools and possibly in other domains.  For instance, we
speculate that some of the heuristics used in
FlashFill (to take one well-documented example) may be connected to some of the principles
laid out here.  For instance, Flash Fill prioritizes the substring constructor over the
constant string constructor when ranking possible programs. Such a choice is
consistent with our information-theoretic viewpoint as the constant function
throws away a great deal of information about the source string being
transformed. Likewise, Flash Fill prefers ``wider'' character classes over
``narrower'' ones. Again such a choice is implied by information theory -- the
wider class preserves more information during translation.  More broadly, we
hope our information theoretic analysis provides a principle that can be
used to understand the heuristic choices made in related work.

@@

More broadly, information theory and probabilistic languages are common tools in
natural language understanding, machine translation, information retrieval, data
extraction and grammatical inference (see Pereira~\cite{pereira:info-theory},
for an introduction). Indeed, our work was inspired, in part, by the PADS format
inference tool~\cite{pads:synthesis}, which was in turn inspired by earlier work
on probabilistic gramatical inference and information
extraction~\cite{kushmerick-thesis,arasu:extracting}. PADS did not synthesize
data transformers, and we are not aware of the use of information-theoretic
principles in type-directed or syntax-guided synthesis of DSL programs.

@@

% begin conclusion
\section{Conclusion}
\label{sec:conc}
We have developed a synthesis algorithm for synthesizing synchronization
functions between data formats that may not be in bijective correspondence. More
specifically, we identify a subset of symmetric lenses, simple symmetric lenses,
develop new combinators for them, and show how to synthesize them from regular
expression specifications. In order to guide the search for ``likely'' lenses,
we introduce new search principles based on information theory. Based on these
principles, we designed, implemented, and evaluated a new tool for lens
synthesis.

@@

\begin{centermath}
\begin{array}[b]{l@{\qquad \; \qquad}l}
  \inferrule*
  {
    \String \in \LanguageOf{\BRegex}\\
    \StringAlt \in \LanguageOf{\BRegexAlt}
  }
  {
    \DisconnectOf{\BRegex}{\BRegexAlt}{\String}{\StringAlt}
    \OfType \BRegex \Leftrightarrow \BRegexAlt
  }
&
  \begin{array}{@{}r@{\ }c@{\ }l@{}}
    $\CreateR{} \App \String''''$ & = & $\StringAlt$\\
    $\CreateL{} \App \StringAlt'''''$ & = & $\String$\\
    $\PutR{} \App \String''' \App \StringAlt'$ & = & $\StringAlt'$\\
    $\PutL{} \App \StringAlt' \App \String'$ & = & $\String'$
  \end{array}
\end{array}
\end{centermath}
